Testing Gravity using Type Ia Supernovae Discovered by LSST

### 2018-11-01 doc/article.tex
forging on

The precision in  $f\sigma_8$ derived from an ten-year LSST-discovered  SN~Ia survey has been projected by \citet{2017ApJ...847..128H},
from both their  clustering and peculiar velocities.
The results are summarized in Table~\ref{tab:howlett}, where a 5\% distance uncertainty is assumed for each supernova.
SNe~Ia alone can provide a $2\%$ measurement of $f\sigma_8$ at $z<0.3$ redshifts lower than where galaxy, cluster, and Ly$\alpha$
RSD measurements are sensitive.

@@

The projected precisions for LSST-discovered SNe~Ia have a number of interesting features.  Particular velocities contribute increasingly to
$f\sigma_8$ precision with decreasing redshift.  At redshift 0.3 the differential gain from adding peculiar velocities to the RSD measurement is small.
The differential gain in the cumulative $f\sigma_8$ constraint is small.  We therefore conclude that the focus should be in $z<0.3$ LSST SN~Ia discoveries.
After 10 years, the total number density of supernovae is such that both sample and shot noise contribute to the error budget for the volume with $z<0.2$, 
Galaxy RSD plus SN~Ia peculiar velocities have yet to be calculated.

They perform their analysis in Fourier space, 
where the Fisher information matrix of a random Gaussian field with mean zero and covariance $C(k)$ parameterized by $\lambda$ is
approximated by
\begin{equation}
F_{ij} = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right],
\end{equation}
where the he covariance
\begin{equation}
C = P_{vv}(k) + \frac{\sigma^2}{n}
\end{equation}
has contributions from the power spectrum, noise in the velocity measurement, and the density of velocity probes
\citep{2017MNRAS.464.2517H}. 
In the sample variance limit for a sample with fixed depth, the variance in $f\sigma_8$ (and other $\lambda$ parameters)
is thus inversely proportional to the survey solid-angle $\Omega$, whereas
in the shot-noise limit the variance is inversely proportional to $\Omega n^2 \propto N^2/\Omega$, where $n$ is the number density,
and $N$ is the total number of supernovae.

While there have been a number of articles on the subject,
our analysis brings a higher level of fidelity than sought by previous analyses.  We simulate SNe~Ia hosted by galaxies in a mock galaxy
catalog. The numbers of SNe are sufficiently small to allow fast evaluations of the likelihood, which enable the determination of parameter
posteriors using MCMC on reasonable computing timescales.   We can use our machinery to 
compare different survey parameters, such as redshift depth, total numbers of supernovae,
solid angle/survey geometry, and SN~Ia intrinsic magnitude dispersion.

The motivation of this work is to quantify the probative power of SN~Ia-derived peculiar velocities in the LSST era.

We consider a maximum redshift of $z=0.2$ because over ten years the noise is not shot-noise dominated, meaning that the differential improvement
in the precision of
$f\sigma_8$ due to a lower-redshift supernova is greater than that due to a high-redshift supernova.  Following all LSST SNe~Ia below this redshift 
likely saturate available follow-up resources.
Our calculations are thus based on the number and solid-angle of $z<0.2$ supernova pre-maximum discoveries 
for the survey candidates provided by the Project.  Based on these numbers we calculate the Figures of Merit in
both sample and shot-noise limits, normalized to ``baseline2018a''.
After 10 years,
LSST supernovae are at neither extreme; we thus adopt the average of the two FoM's as what we advocate for the survey.
These averages for a select set of surveys are shown in Figure~\ref{fig:fom}.

The cross-correlation between galaxy-count and peculiar-velocity surveys is not yet implemented but is a planned extension of this work.  We would like to
quantify the suppression of sample variance achieved when considering matter-densities and velocities 
within the same volume \citep{2007PhRvL..99h1301G}.

\section{Simulated Data}
The Buzzard (v1.6) galaxy catalog is used.  This is because it is the survey, with a light cone that covers 10,313.24 sq.~deg., that covers
the largest solid angle among those currently available
in the DESC Generic Catalog Reader.  However,
the survey geometry is different
from that of LSST.  The catalog is based on a Flat $\Lambda$CDM model with $H_0=70$~km~s$^{-1}$,  $\Omega_M=0.286$, $\Omega_B=0.047$, and
$\Omega_\nu=0$.
The catalog contains 140M galaxies with observed redshift $z<0.2$.
Each galaxy has its cosmological redshift, the $x$-, $y$-, $z$-components of its peculiar velocity from which
the radial peculiar velocity is determined, and its
star formation rate and stellar mass from which the
supernova rate of each host is determined using 
\citet{2012ApJ...755...61S}.  Supernovae are realized for a span of 10 observer-years yielding 17,205 objects.  This total supernova
production underestimates the expected discovery of 130,000 supernovae in 10 observer-years based on the volumetric
rate of \citet{2010ApJ...713.1026D}.  The volumetric rate is more robust, so the sample under consideration in this article
corresponds to 1.3 years worth of discoveries.

@@

@@

\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{|cccc|cccc|} % Column formatting, @{} suppresses leading/trailing space
   \hline
$z_{max}$ & fraction & $\sigma_{SN}$ & $N_{gal}$ & $\bar{A}$ & $\sigma_A$ & $\sigma_A N_{gal}^{-0.5}$ & $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$ \\
\hline
0.20 &  1.0 & 0.08 & 9901 &   1.18 &   0.23 &  0.050 &  24.38 \\ \hline
0.20 &  0.2 & 0.08 & 2021 &   1.66 &   0.59 &  0.062 &  13.61 \\
0.20 &  0.3 & 0.08 & 3023 &   1.34 &   0.43 &  0.057 &  15.16 \\
0.20 &  0.5 & 0.08 & 5015 &   1.13 &   0.30 &  0.054 &  18.54 \\
0.10 &  1.0 & 0.08 & 1294 &   0.91 &   0.28 &  0.091 &  15.96 \\
0.15 &  1.0 & 0.08 & 3870 &   1.00 &   0.24 &  0.066 &  20.00 \\
0.20 &  1.0 & 0.10 & 9901 &   1.06 &   0.25 &  0.042 &  20.33 \\
0.20 &  1.0 & 0.12 & 9901 &   1.09 &   0.28 &  0.040 &  19.13 \\
0.15 &  0.3 & 0.12 & 1215 &   1.97 &   0.92 &  0.061 &  10.36 \\
0.12 &  0.3 & 0.10 & 746 &   1.62 &   0.70 &  0.085 &  11.28 \\
0.07 &  1.0 & 0.08 & 563 &   1.16 &   0.36 &  0.135 &  15.55 \\
    \hline
   \end{tabular}
   \caption{.}
   \label{tab:booktabs}
\end{table}

>>>>

The precision in  $f\sigma_8$ derived from an ten-year LSST-discovered  SN~Ia survey has been projected by \citet{2017ApJ...847..128H},
from both their  clustering and peculiar velocities.
The results are summarized in Table~\ref{tab:howlett}, where a 5\% distance uncertainty is assumed for each supernova.
SNe~Ia alone can provide a $2\%$ measurement of $f\sigma_8$ at $z<0.3$ redshifts lower than where galaxy, cluster, and Ly$\alpha$
RSD measurements are sensitive.
{\bf Greg do you have to power to add a SN PV + DESI-like BGS column and/or a PV-only column with Cullan's code? }

@@

The projected precisions for LSST-discovered SNe~Ia have a number of interesting features.  Particular velocities contribute increasingly to
$f\sigma_8$ precision with decreasing redshift.  At redshift 0.3 the differential gain from adding peculiar velocities to the RSD measurement is small.
The differential gain in the cumulative $f\sigma_8$ constraint is small.  We therefore conclude that the focus should be in $z<0.3$ LSST SN~Ia discoveries.
After 10 years for $z>0.1$ both sample variance and shot noise contribute to the error budget; adding even more SNe in the volume would improve
the measurement though not as $\sqrt{N}$.

\section{Supernova Peculiar Velocity Survey}
\citet{2017ApJ...847..128H} perform their analysis in Fourier space, 
where the Fisher information matrix of a random Gaussian field with mean zero and covariance $C(k)$ parameterized by $\lambda$ is
approximated by
\begin{equation}
F_{ij} = \frac{V}{2}\int \frac{d^3k}{(2\pi)^3} \text{Tr}\left[ C^{-1} \frac{\partial C}{\partial \lambda_i} C^{-1}
\frac{\partial C}{\partial \lambda_j} \right],
\end{equation}
where the covariance
\begin{equation}
C = P_{vv}(k) + \frac{\sigma^2}{n}
\end{equation}
has contributions from the power spectrum, noise in the velocity measurement, and the density of velocity probes
\citep{2017MNRAS.464.2517H}.   This equation is for the peculiar-velocity auto-correlation.

The parameters that determine parameter precision are the volume $V$, which
in turn can be parameterized by the survey solid angle $\Omega$ and the redshift depth $z_{max}$ ; the number density of sources $n$; and the intrinsic
magnitude dispersion $\sigma_M$, where magnitude and velocity dispersions are related by $\sigma_{M} = \frac{5}{\ln{10}} \frac{1+z}{z} \sigma$.
In the sample variance limit for a sample with fixed depth, the variance in $f\sigma_8$ (and other $\lambda$ parameters)
is thus inversely proportional to the survey solid-angle $\Omega$, whereas
in the shot-noise limit the variance is inversely proportional to $\Omega n^2 \propto N^2/\Omega$, where $n$ is the number density,
and $N$ is the total number of supernovae.

LSST is expected to discover all $z<0.3$ SNe~Ia before maximum light in its active survey area.

The quality of supernova follow-up is an important factor in how well a peculiar-velocity survey can measure $f\sigma_8$.
LSST light curves of low-redshift SNe, with their relatively poor sampling, plus a spectroscopic redshift
 can give $\sigma_M \sim 0.15$~mag. Non-LSST follow-up including infrared data or spectrophotometry at peak brightness are projected to give
$\sigma_M \lesssim 0.08$~mag.
Uncertainties in $f\sigma_8$ depend on $\sigma^2_M/n$ so the improvement between 0.08 and 0.15~mag dispersions is equivalent to a factor of 3.52
 in number density, or in survey duration given the expected completeness of $z<0.3$ SN~Ia discoveries.

The variance in $f\sigma_8$ is proportional to solid angle $\Omega$ given a fixed redshift depth and  $\sigma^2_M/n$.  The baseline
LSST survey covers 18,000 sq.~deg, at $-75 \lesssim \delta \lesssim15$ avoiding the Galactic plane.  Larger solid-angle coverage for $z<-0.3$ SN~Ia searches, beyond
the LSST baseline, benefits peculiar-velocity science.  Complementary northern-hemisphere surveys and LSST-expanded or independent
coverage of the southern pole, could double
the sky coverage and halve the variance in $f\sigma_8$.

LSST can be taken to be a SN~Ia discovery plus light-curve builder, or a discovery machine only.  The former comes with a simplified follow-up program of typing
and/or redshift collection but with higher intrinsic magnitude dispersion.  The latter requires additional follow-up resources to gather more information per supernova to
lower the intrinsic magnitude dispersion.

\appendix
\section{Reanalysis in Configuration Space}

This Appendix describes a new analysis used to project precisions on $f\sigma_8$ using peculiar velocities derived from SNe~Ia.
While there have been a number of articles on the subject,
our analysis brings a higher level of fidelity than sought by previous analyses.  We simulate SNe~Ia hosted by galaxies in a mock galaxy
catalog. The numbers of SNe are sufficiently small to allow fast evaluations of the likelihood, which enable the determination of parameter
posteriors using MCMC on reasonable computing timescales.   We can use our machinery to 
compare different survey parameters, such as redshift depth, total numbers of supernovae,
solid angle/survey geometry, and SN~Ia intrinsic magnitude dispersion.
We would like to
quantify the suppression of sample variance achieved when considering matter-densities and velocities 
within the same volume \citep{2007PhRvL..99h1301G}.
The cross-correlation between galaxy-count and peculiar-velocity surveys is not yet implemented but is a planned extension of this work.

\section{Simulated Data}
The cosmoDC2 (v1.0) covering 706 sq.~deg is used\footnote{We previously used
Buzzard (v1.6) galaxy catalog is used, but it was found to give peculiar velocities inconsistent with general relativity.  As we will see,
cosmoDC2 (v1.0) is consistent with GR. }. 
The catalog is based on a Flat $\Lambda$CDM model with $H_0=71$~km~s$^{-1}$,  $\Omega_M=0.265$, $\Omega_B=0.0448$, and
$\Omega_\nu=0$.
Each galaxy has its cosmological redshift, the $x$-, $y$-, $z$-components of its peculiar velocity from which
the radial peculiar velocity is determined, and its
star formation rate and stellar mass from which the
supernova rate of each host is determined using 
\citet{2012ApJ...755...61S}.   This host-based rate
underestimates the expected discovery based on the volumetric
rate of \citet{2010ApJ...713.1026D}.  The volumetric rate is more robust, so  we boost the SN production
by a factor 2.4 to get the 10-year volumetric rate expectation. The total number of simulated SNe 
does overestimate the total number of pre-maximum supernova discoveries, since most fields cannot continuously monitored throughout the year
from most ground-based observatories.

@@

We consider a maximum redshift of $z=0.2$ because over ten years the noise is not shot-noise dominated, meaning that the differential improvement
in the precision of
$f\sigma_8$ due to a lower-redshift supernova is greater than that due to a high-redshift supernova.  Following all LSST SNe~Ia below this redshift 
would likely saturate available follow-up resources.
Our calculations are thus based on the number and solid-angle of $z<0.2$ supernova pre-maximum discoveries 
for the survey candidates provided by the Project.

@@

\begin{table}
   \centering
   %\topcaption{Table captions are better up top} % requires the topcapt package
   \begin{tabular}{|ccc|ccccc|} % Column formatting, @{} suppresses leading/trailing space
   \hline
$z_{max}$ & fraction & $\sigma_{SN}$ & $N_{gal}$ & $\bar{A}$ & $\sigma_A$ & $\sigma_A N_{gal}^{-0.5}$ & $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$ \\
\hline
0.20 & 0.20 & 0.08 & 2021 &   1.66 &   0.59 &  0.062 &  13.61 \\
0.10 & 0.30 & 0.08 & 429 &   1.26 &   0.51 &  0.120 &  12.09 \\
0.15 & 0.30 & 0.08 & 1215 &   1.42 &   0.55 &  0.074 &  12.48 \\
0.15 & 0.30 & 0.12 & 1215 &   1.97 &   0.92 &  0.061 &  10.36 \\
0.20 & 0.30 & 0.08 & 3023 &   1.34 &   0.43 &  0.057 &  15.16 \\
0.20 & 0.30 & 0.10 & 3023 &   1.40 &   0.54 &  0.047 &  12.68 \\
0.20 & 0.30 & 0.12 & 3023 &   1.48 &   0.72 &  0.037 &   9.95 \\
0.15 & 0.50 & 0.08 & 1982 &   1.12 &   0.37 &  0.069 &  14.93 \\
0.20 & 0.50 & 0.08 & 5015 &   1.13 &   0.30 &  0.054 &  18.54 \\
0.07 & 0.65 & 0.08 & 392 &   1.34 &   0.49 &  0.137 &  13.21 \\
0.10 & 0.65 & 0.08 & 873 &   0.89 &   0.35 &  0.087 &  12.57 \\
0.10 & 0.65 & 0.10 & 873 &   1.01 &   0.46 &  0.075 &  10.72 \\
0.15 & 0.65 & 0.08 & 2541 &   1.15 &   0.33 &  0.070 &  17.05 \\
0.15 & 0.65 & 0.10 & 2541 &   1.15 &   0.39 &  0.059 &  14.43 \\
0.15 & 0.65 & 0.12 & 2541 &   1.30 &   0.51 &  0.051 &  12.42 \\
0.20 & 0.65 & 0.08 & 6491 &   1.28 &   0.30 &  0.053 &  20.79 \\
0.20 & 0.65 & 0.10 & 6491 &   1.24 &   0.34 &  0.046 &  17.95 \\
0.20 & 0.65 & 0.12 & 6491 &   1.33 &   0.45 &  0.037 &  14.33 \\
0.07 & 1.00 & 0.08 & 563 &   1.16 &   0.36 &  0.135 &  15.55 \\
0.10 & 1.00 & 0.08 & 1294 &   0.91 &   0.28 &  0.091 &  15.96 \\
0.15 & 1.00 & 0.08 & 3870 &   1.00 &   0.24 &  0.066 &  20.00 \\
0.20 & 1.00 & 0.08 & 9901 &   1.18 &   0.23 &  0.050 &  24.38 \\
0.20 & 1.00 & 0.10 & 9901 &   1.06 &   0.25 &  0.042 &  20.33 \\
0.20 & 1.00 & 0.12 & 9901 &   1.09 &   0.28 &  0.040 &  19.13 \\
    \hline
   \end{tabular}
   \caption{Signal-to-noise in $A$ or equivalently $f\sigma_8$ for surveys parameterized by the maximum redshift $z_{max}$,
   the fraction of total supernovae in 10 years ``fraction'', intrinsic magnitude dispersion $\sigma_{SN}$.  The number of galaxies
   $N_{gal}$, the signal $\bar{A}$ and its uncertainty $\sigma_A$ are for the cosmoDC2 (v1.0) 760 sq.~deg. volume only.
   The uncertainty ``per'' supernova is given by  $\sigma_A N_{gal}^{-0.5}$.  The signal-to-noise, scaled to an LSST solid angle, is given as 
    $\bar{A} \sigma_A^{-1} (18000/760)^{0.5}$.
   \label{tab:booktabs}}
\end{table}
